Legal Analytics Problem Set 3 - Part 2
-----------------------------------------------------------------------------
## Andy Ninh
-----------------------------------------------------------------------------

## Loading the Gapminder data
```{r}
gdURL <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt"
gDat <- read.delim(file = gdURL)
```

```{r}
str(gDat)
```

## Storing a snippet of the data
```{r}
(snippet <- subset(gDat, country == "Canada"))
```

## Installing and Loading plyr
```{r}
install.packages("plyr", dependencies = TRUE)
library(plyr)
```

## Using ddply to find the maximum life expentency of each continent
```{r}
(maxLeByCont <- ddply(gDat, ~ continent, summarize, maxLifeExp = max(lifeExp)))
```

## Studying the return value
```{r}
str(maxLeByCont)
levels(maxLeByCont$continent)
```

## Computing the minimum GDP per capita by continent
```{r}
(mingdpPercap <- ddply(gDat, ~ continent, summarize, minGDPperCap = min(gdpPercap)))
```

## Counting the number of countries in the data set
```{r}
ddply(gDat, ~continent, summarize, nUniqCountries = length(unique(country)))
```

## Alternate method without using summarize()
```{r}
ddply(gDat, ~ continent,
      function(x) return(c(nUniqCountries = length(unique(x$country)))))
```

## You can compute multiple things per sub-data.frame
```{r}
ddply(gDat, ~ continent, summarize,
      minLifeExp = min(lifeExp), maxLifeExp = max(lifeExp),
      medGdpPercap = median(gdpPercap))
```

# Putting it all together using ddply()
```{r}
jCountry <- "France"  # pick, but do not hard wire, an example
(jDat <- subset(gDat, country == jCountry))  # temporary measure!
```

## Installing and loading lattice for plotting graphics
```{r}
install.packages("lattice")
library(lattice)
```

## Plotting the data 
```{r}
xyplot(lifeExp ~ year, jDat, type = c("p", "r"))  # always plot the data
jFit <- lm(lifeExp ~ year, jDat)
summary(jFit)
```

## Adjusting for fit

```{r}
(yearMin <- min(gDat$year))
jFit <- lm(lifeExp ~ I(year - yearMin), jDat)
summary(jFit)
## The intercept is around 68. This makes much more sense. So, what is jFit and what can we get out of it?
class(jFit)
mode(jFit)
## So, jFit is of class "lm" and its mode is list. So that means I could use indexing to isolate specific components. But what's in there?
```

```{r}
## str(jFit) # too ugly to print here but you should look
names(jFit)
jFit$coefficients
## Using str() and names() reveals a great deal about this "lm" object and reading the help file for lm() would explain a great deal more. In the See Also section we learn there's a generic function coef() which looks promising.
coef(jFit)
## We have achieved our goal for this specific country -- we've gotten its intercept and slope. Now we need to package that as a function.
jFun <- function(x) coef(lm(lifeExp ~ I(year - yearMin), x))
jFun(jDat)  # trying out our new function ... yes still get same numbers
```

## Changing the names of the return values
```{r}
jFun <- function(x) {
    estCoefs <- coef(lm(lifeExp ~ I(year - yearMin), x))
    names(estCoefs) <- c("intercept", "slope")
    return(estCoefs)
}
jFun(jDat)  # trying out our improved function ... yes still get same numbers
```

## Testing the new function on some examples
```{r}
jFun(subset(gDat, country == "Canada"))
jFun(subset(gDat, country == "Uruguay"))
jFun(subset(gDat, country == "India"))
jFun(subset(gDat, country == "Japan"))
jFun(subset(gDat, country == "Iceland"))
```

## Scaling up -- Placing the function inside ddply()
```{r}
jCoefs <- ddply(gDat, ~country, jFun)
str(jCoefs)
tail(jCoefs)
```

# Review: This would be the final script
```{r}
## realistically, you would read the data from a local file
gdURL <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt"
gDat <- read.delim(file = gdURL)
## str(gDat) here when working interactively
yearMin <- min(gDat$year)
jFun <- function(x) {
    estCoefs <- coef(lm(lifeExp ~ I(year - yearMin), x))
    names(estCoefs) <- c("intercept", "slope")
    return(estCoefs)
}
## jFun(subset(gDat, country == 'India')) to see what it does
jCoefs <- ddply(gDat, ~country, jFun)
```
## Installing and loading xtable 
```{r}
install.packages("xtable", dependencies = TRUE)
library(xtable)
```

## Pick some countries at random and display their estimated coefficients
```{r results='asis'}
set.seed(916)
foo <- jCoefs[sample(nrow(jCoefs), size = 15), ]
foo <- xtable(foo)
print(foo, type = "html", include.rownames = FALSE)
```
## Two easy improvements: include the continent informaiton and sort rationally
```{r}
jCoefs <- ddply(gDat, ~country + continent, jFun)
str(jCoefs)
tail(jCoefs)
```
## Now, prior to making the HTML table, we will sort the data.frame, so it starts with the country with the shortest life expectancy in 1952, and goes to the largest.
```{r results='asis'}
set.seed(916)
foo <- jCoefs[sample(nrow(jCoefs), size = 15), ]
foo <- arrange(foo, intercept)
## foo <- foo[order(foo$intercept), ] # an uglier non-plyr way
foo <- xtable(foo)
print(foo, type = "html", results='asis', include.rownames = FALSE)
```

# jFun only requires one argument, x. What if it had more than one argument?
```{r}
jFunTwoArgs <- function(x, cvShift = 0) {
    estCoefs <- coef(lm(lifeExp ~ I(year - cvShift), x))
    names(estCoefs) <- c("intercept", "slope")
    return(estCoefs)
}
jCoefsSilly <- ddply(gDat, ~country, jFunTwoArgs)
head(jCoefsSilly)
## We are getting the same estimated slopes but the silly year 0 intercepts we've seen before. Let's use the cvShift = argument to resolve this.
jCoefsSane <- ddply(gDat, ~country, jFunTwoArgs, cvShift = 1952)
head(jCoefsSane)
## We're back to our usual estimated intercepts, which reflect life expectancy in 1952. Of course hard-wiring 1952 is not a great idea, so here's probably our best code yet:
jCoefsBest <- ddply(gDat, ~country, jFunTwoArgs, cvShift = min(gDat$year))
head(jCoefsBest)
```
